---
title: "The Relationship between Body Size and Lifespan"
author: "Juan Vazquez"
date: 2017-04-21
output: html_document
---

## Introduction

While aging is an inevitable process for most species, there is an incredible diversity
of lifespans throughout the Tree of Life, ranging from a few days to [several millenia][https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4133289/]. For researchers interested
in the fundamental biology behind ageing, seeing what aspects of an organism's biology
correlate to lifespan is an important first step on the path to finding concrete explanations 
behind their longevity.  
For example, in 1975, Dr. Richard Peto published [a paper][https://www.ncbi.nlm.nih.gov/pubmed/1212409]
where he established that the different sizes and lifespans of humans and mice didn't really 
relate to their respective cancer rates. This was described as Peto's Paradox, because 
the expectation was originally that over a lifetime, every cell will accumulate mutations that 
could eventually cause it to become cancerous; and if an animal had more cells, then this 
lifetime risk of cancer would only increase further. In fact, it turns out that there 
is no relationship between body size, lifespan, and cancer, which is the fact that 
underlies the focus of my own research!  
As we will explore in this section, this paradox is further complicated by another 
unexpected relationship: animals that are larger tend to also live longer.

## The Data

For this analysis, we will be using the [AnAge database][] of ageing and life history 
in animals. This database has entries for over 4200 species of animals 
(also 2 plants and 3 fungi) with data like max lifespan, growth rates and weights at
different life stages, descriptions, and metabolism, amongst other things.  

First, let's take a look at the data itself:

```{r packages, message=FALSE}
# These are the packages we will be using in this analysis
require(dplyr)
require(ggplot2)
require(plotly)
```

```{r load_data, message=FALSE}
# Read the data into a dataframe:
anage <- read.delim('../data/anage_data.txt',as.is = T)
# Look at the data using str()
str(anage)
```

Using str() is a useful way of quickly seeing the different columns of data 
and the type of data in each. Of interest are the taxonomic columns, the  
Adult Weight column, and the Maximum Lifespan column.
You can also try use head() to look at the first few rows:  

```{r see_data}
head(anage)
```

As you can see, head is less legible in dataframes with many columns, so I usually stick to str().  
  
Let's try graphing now. First, we will graph all the species; we 
will graph the adult weights versus their maximum lifespan, and color the 
datapoints by their Phylum:  


```{r plot_all}
# Create the basic plot
p_all <- ggplot(
  data = anage,
  aes(Adult.weight..g., Maximum.longevity..yrs., color=Phylum)
  )
# Add the features
p_all <- p_all + 
  geom_point(aes(text = paste(Common.name)), size=0.5) + 
  scale_x_log10() + 
  ggtitle('AnAge Species Adult Weight vs Lifespan') + 
  xlab('Adult Weight (log(g))') + 
  ylab('Maximum Longevity (yrs)')
# Output the interactive plot
ggplotly(p_all)
```

*(Note that we scaled the x-axis using a log-scale; this is done because we want 
to highlight orders-of-magnitude changes over small-scale change - in other words,
we don't care so much about the difference between 1-2 grams and 100-200 grams as
much as a change between 1-10 grams and 100-1000 grams.)*  

As you can see, the graph looks a little off! It seems like a straight line, 
but that's because the Y-axis is extremely shifted; there's no point anywhere
near that high. We can look back at the data using dplyr to try to figure 
out why the plot was set so high:  

```{r get_outliers}
# Filter anage based on the weird characteristics:
anage %>% filter(Maximum.longevity..yrs. >= 1000)
```
Remember how I said that some animals live for millenia? Behold the humble sponge, 
and specifically the Hexactinellid sponge, which apparently lives for 15,000 years! 
Its worth noting the column "Data.quality" here; there's some skepticism in the 
literature as to whether or not this is estimate is real, since its so incredible. 
Also of note is why these points don't appear on the graph - none of them have 
data for adult weight! Let's try again, but this time we'll filter out species
without entries for weight or lifespan:  


```{r plot_all2}
# Filter out missing values:
anage %>% filter(!is.na(Adult.weight..g.), !is.na(Maximum.longevity..yrs.)) -> anage_filtered
# Create the basic plot
p_all2 <- ggplot(
  data = anage_filtered,
  aes(Adult.weight..g., Maximum.longevity..yrs., color=Phylum)
  )
# Add the features
p_all2 <- p_all2 + 
  geom_point(aes(text = paste(Common.name)), size=0.5) + 
  scale_x_log10() + 
  ggtitle('AnAge Species Adult Weight vs Lifespan') + 
  xlab('Adult Weight (log(g))') + 
  ylab('Maximum Longevity (yrs)')
# Output the interactive plot
ggplotly(p_all2)
```

Next, let us graph all chordates:

```{r plot_chordates}
anage %>% filter(Phylum == 'Chordata') -> anage_chordates
p_chordates <- ggplot(
  data = anage_chordates,
  aes(Adult.weight..g., Maximum.longevity..yrs., color=Class)
  )
p_chordates <- p_chordates + geom_point(aes(text = paste(Common.name)), size=0.5) + scale_x_log10() + ggtitle('AnAge Species Adult Weight vs Lifespan') + xlab('Adult Weight (log(g))') + ylab('Maximum Longevity (yrs)')
ggplotly(p_chordates)

```

Finally, let us look at bats:

```{r plot_bats}

anage %>% filter(Order=='Chiroptera') -> anage_chiroptera
str(anage_chiroptera)

p_bats <- ggplot(
  data = anage_chiroptera,
  aes(Adult.weight..g., Maximum.longevity..yrs., color=Family)
)
p_bats <- p_bats + geom_point(aes(text = paste(Common.name))) + scale_x_log10() + ggtitle('Chiroptera Adult Weight vs Lifespan') + xlab('Adult Weight (log(g))') + ylab('Maximum Longevity (yrs)')
ggplotly(p_bats)
```


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

[AnAge database]: http://genomics.senescence.info/species/
