---
title: "The Relationship between Body Size and Lifespan"
author: "Juan Vazquez"
date: 2017-04-21
output: html_document
---

## Introduction

While aging is an inevitable process for most species, there is an incredible diversity
of lifespans throughout the Tree of Life, ranging from a few days to [several millenia][]. For researchers interested
in the fundamental biology behind ageing, seeing what aspects of an organism's biology
correlate to lifespan is an important first step on the path to finding concrete explanations 
behind their longevity.  
For example, in 1975, Dr. Richard Peto published [a paper][]
where he established that the different sizes and lifespans of humans and mice didn't really 
relate to their respective cancer rates. This was described as Peto's Paradox, because 
the expectation was originally that over a lifetime, every cell will accumulate mutations that 
could eventually cause it to become cancerous; and if an animal had more cells, then this 
lifetime risk of cancer would only increase further. In fact, it turns out that there 
is no relationship between body size, lifespan, and cancer, which is the fact that 
underlies the focus of my own research!  
As we will explore in this section, this paradox is further complicated by another 
unexpected relationship: animals that are larger tend to also live longer.

## Graphing the Data

For this analysis, we will be using the [AnAge database][] of ageing and life history 
in animals. This database has entries for over 4200 species of animals 
(also 2 plants and 3 fungi) with data like max lifespan, growth rates and weights at
different life stages, descriptions, and metabolism, amongst other things.  

First, let's take a look at the data itself:

```{r packages, message=FALSE}
# These are the packages we will be using in this analysis
require(dplyr)
require(ggplot2)
require(plotly)
```

```{r load_data, message=FALSE}
# Read the data into a dataframe:
anage <- read.delim('../data/anage_data.txt',as.is = T)
# Look at the data using str()
str(anage)
```

Using str() is a useful way of quickly seeing the different columns of data 
and the type of data in each. Of interest are the taxonomic columns, the  
Adult Weight column, and the Maximum Lifespan column.
You can also try use head() to look at the first few rows:  

```{r see_data}
head(anage)
```

As you can see, head is less legible in dataframes with many columns, so I usually stick to str().  
  
### All species 

Let's try graphing now. First, we will graph all the species; we 
will graph the adult weights versus their maximum lifespan, and color the 
datapoints by their Phylum:  


```{r plot_all}
# Create the basic plot
p_all <- ggplot(
  data = anage,
  aes(Adult.weight..g., Maximum.longevity..yrs., color=Phylum)
  )
# Add the features
p_all <- p_all + 
  geom_point(aes(text = paste(Common.name)), size=0.5) + 
  scale_x_log10() + 
  ggtitle('AnAge Species Adult Weight vs Lifespan') + 
  xlab('Adult Weight (log(g))') + 
  ylab('Maximum Longevity (yrs)')
# Output the interactive plot
ggplotly(p_all)
```

*(Note that we scaled the x-axis using a log-scale; this is done because we want 
to highlight orders-of-magnitude changes over small-scale change - in other words,
we don't care so much about the difference between 1-2 grams and 100-200 grams as
much as a change between 1-10 grams and 100-1000 grams.)*  

As you can see, the graph looks a little off! It seems like a straight line, 
but that's because the Y-axis is extremely shifted; there's no point anywhere
near that high. We can look back at the data using dplyr to try to figure 
out why the plot was set so high:  

```{r get_outliers}
# Filter anage based on the weird characteristics:
anage %>% filter(Maximum.longevity..yrs. >= 1000)
```
Remember how I said that some animals live for millenia? Behold the humble sponge, 
and specifically the Hexactinellid sponge, which apparently lives for 15,000 years! 
Its worth noting the column "Data.quality" here; there's some skepticism in the 
literature as to whether or not this is estimate is real, since its so incredible. 
Also of note is why these points don't appear on the graph - none of them have 
data for adult weight! Let's try again, but this time we'll filter out species
without entries for weight or lifespan:  


```{r plot_all2}
# Filter out missing values:
anage %>% filter(!is.na(Adult.weight..g.), !is.na(Maximum.longevity..yrs.)) -> anage_filtered
# Create the basic plot
p_all2 <- ggplot(
  data = anage_filtered,
  aes(Adult.weight..g., Maximum.longevity..yrs., color=Phylum)
  )
# Add the features
p_all2 <- p_all2 + 
  geom_point(aes(text = paste(Common.name)), size=0.5) + 
  scale_x_log10() + 
  ggtitle('AnAge Species Adult Weight vs Lifespan, Filtered') + 
  xlab('Adult Weight (log(g))') + 
  ylab('Maximum Longevity (yrs)')
# Output the interactive plot
ggplotly(p_all2)
```

Interestingly, it seems like all non-chordates lack one or the other datapoint...  

### Chordates 
Moving on, let us graph the chordates, and now color by class:  

```{r plot_chordates}
# Filter the database so that we only have species from Chordata,
#  with both adult weight and maximum longevity.
anage %>% filter(Phylum == 'Chordata',!is.na(Adult.weight..g.), !is.na(Maximum.longevity..yrs.)) -> anage_chordates
# Basic Graph
p_chordates <- ggplot(
  data = anage_chordates,
  aes(Adult.weight..g., Maximum.longevity..yrs., color=Class)
  )
# Fun stuff
p_chordates <- p_chordates + 
  geom_point(aes(text = paste(Common.name)), size=0.5) + 
  scale_x_log10() + 
  ggtitle('Chordates Adult Weight vs Lifespan') + 
  xlab('Adult Weight (log(g))') + 
  ylab('Maximum Longevity (yrs)')
# And plot
ggplotly(p_chordates)

```

Overall, there seems to be a positive trend; we can already see the outliers
that are extremely long-lived for their size by hovering over the points 
to read the common names. As you can see, the longest lived animals by far 
are the Rougheye Rockfish, at 205 years; and the Bowhead Whale, at 211 years.
A little further down the y axis, we see many more fish, humans (!), and some more whales. 
Now we can look further at the data by clicking on the classes on the legend 
to hide those points.  
  
Hide everything so that we can only see Actinopterygii, which is the class of 
ray-finned fish. This group accounts for almost all of the living fish species, and
makes up half of all living vertebrates. As you can see, for any given weight 
of fish, you could probably find many short lived fish and a couple of really 
long-lived fish. In contrast, if you look only at Amphibians, you can see that
based on our data they pretty much cluster tightly with the exception of the
Olm, which lives a very long time for its size.  
  
Looking at Aves, or birds, we can see that there's actually a fairly clear 
positive relationship between size and longevity, with most of the density 
in the plot at the small-and-short-lived end of the tail, fanning out as you 
get larger and larger. Were I to venture a guess, this fan-shaped pattern 
is suggestive of the two sides of Peto's Paradox: some of these birds seem to 
live longer as per the pattern, but some of them seem to be shorter-lived than
one would expect. One thing that caught my attention was how emus and cassowaries,
which are sister species, are all shorter-lived than the ostrich, which is at 
the root of the Ratite clade. While this alone means nothing, its small things 
of note like this that could lead to the start of some really interesting work!  

### Mammals

Now let us move on to mammals:  

```{r plot_mammals}
# Select all mammals with values for size and lifespan
anage %>% filter(Class == 'Mammalia',!is.na(Adult.weight..g.), !is.na(Maximum.longevity..yrs.)) -> anage_mammalia

# Graph!
p_mammalia <- ggplot(
  data = anage_mammalia,
  aes(Adult.weight..g., Maximum.longevity..yrs., color=Order)
)
p_mammalia <- p_mammalia + 
  geom_point(aes(text = paste(Common.name)), size=0.5) + 
  scale_x_log10() + 
  ggtitle('Mammals Adult Weight vs Lifespan') + 
  xlab('Adult Weight (log(g))') + 
  ylab('Maximum Longevity (yrs)')
ggplotly(p_mammalia)
```

Here we see the beautiful correlation between mammalian weight and lifespan! 
To really get an in-depth look at things, let's add on a few more things:

- First, let's log-transform the y axis as well;  
- Second, we'll use the 'lm' method of geom_smooth() to add a regression line;
- Finally, we'll add some confidence intervals.

And so we get this:  

```{r plot_mammals2}
# The hard work's done so let's just add on the new things
p_mammalia2 <- p_mammalia + 
  scale_y_log10() +
  geom_smooth(method='lm', aes(Adult.weight..g., Maximum.longevity..yrs.), inherit.aes = FALSE)
  
ggplotly(p_mammalia2)
```

Now that we can see the regression line, we can also look to see families that are above the trend line, versus those that are below it. 



### Bats

Finally, let us look at bats:

```{r plot_bats}

anage %>% filter(Order=='Chiroptera',!is.na(Adult.weight..g.), !is.na(Maximum.longevity..yrs.)) -> anage_chiroptera
str(anage_chiroptera)

# Once more, with vigor!
p_bats <- ggplot(
  data = anage_chiroptera,
  aes(Adult.weight..g., Maximum.longevity..yrs., color=Family)
)
p_bats <- p_bats + geom_point(aes(text = paste(Common.name))) + scale_x_log10() + ggtitle('Chiroptera Adult Weight vs Lifespan') + xlab('Adult Weight (log(g))') + ylab('Maximum Longevity (yrs)')
ggplotly(p_bats)
```


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```


[several millenia]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4133289/
[a paper]: https://www.ncbi.nlm.nih.gov/pubmed/1212409
[AnAge database]: http://genomics.senescence.info/species/
